---
title: "Lab 6 Income as Dependent"
author: "Kyle Landolt"
date: "April 11, 2015"
output: html_document
---


# Homework
Your homework assignment is to analyze the prestige data, using "income" as the dependent variable.  That is, develop a regression model that helps to explain the income in an occupation, using any of the other data in the the Prestige data set as a potential predictor.  Describe your methodology from beginning to end.  I have done this using Prestige as the dependent variable, and you should follow along with that before continuing.

```{r}
library(spdep)
library(dplyr)
library(ggplot2)
library(GGally)
library(car)
data(Prestige)
```

```{r}
ggpairs(Prestige %>% filter(!is.na(type)) %>% select(education, prestige, women, type, income))
```

I started to analyze the relationship of income as a dependent factor by using the ggpairs function to quickly observe the bivariate relationship of income to the other independent factors. I note the following:

 1. If the solid, smoothed black line of income represents its distribution, then I notice that it seems to be positively skewed. 
 2. There doesn't seem to a good linear relationship of income to any covariates, although the correlation coefficient for income to prestige was fairly high at +0.703. Education has a relatively high correlation coefficient too at 0.574. It's also noted that income to women have a negative correlation coefficent, possibly showing that women are paid less than males.
 3. The relationships of income to education and prestige both appear exponential.


I do not want to remove any covariates at the moment for risk of creating a confounding variable, so I will do a multivariate regression to see the results.
```{r}
reg0 <- lm(income ~ type + education + prestige + women, data=Prestige)
summary(reg0)
```

The initial multivariate regression has prestige and women as significant relationships. I can check for lineary using the component + residual plots.
```{r}
crPlots(reg0)
```

After checking the plots I noticed that it's still clear that education and prestige have an exponential relationship. Because I don't want to transform my dependent variable first, I will first try to transform education and prestige to exp(education) and exp(prestige). I do this so I won't have to directly transform income in hopes of maintaning my significant relationship of income to women. 
```{r}
reg1 <- lm(income ~ type + exp(education) + exp(prestige) + women, data=Prestige)
summary(reg1)
crPlots(reg1)
```

The summary of this regression shows all the regressed independent variables as significant with women as the only negative coefficient. Looking at the component + residual plots, the exp(prestige) independent variable does not express the linear relationship so desired. The Adjusted R^2 value is also only 0.6521.  I will now only log(income) and observe the results:

```{r}
reg2 <- lm(log(income) ~ type + education + prestige + women, data=Prestige)
summary(reg2)
crPlots(reg2)
```

Because we logged the dependent variable, income, now the regression is interpreted as 1 unit change in x is now 100 $beta$ % change in y. 

After logging income, I see that there are significant relationships with prestige and women and the R^2 value has become 0.7842. The component residual lines are more satisfactory, although I'm still not completely happy with education because of the inconsistency towards the right tail.

I would to now see with the reg2 model how education will change according to a prediction of log(income) holding the values of prestige and women at a constant.
```{r}
# First, copy the data, but then fix eduation and income to set values
pred.data <- Prestige %>% mutate(prestige=50, women=60)
# Now, do the predictions
pred.data$income.predict2 <- predict(reg2, newdata = pred.data)
# And plot
ggplot(pred.data) + geom_point(aes(x=education, y=income.predict2, color=type))+
  labs(title = "Prediction with log(income) = a + b education" )
```

You can clearly see how education and the predicted log income relates to types. However, I will now try to run a linear regression excluding education to see how that will affect the adjusted R^2 value.
```{r}
reg3 <- lm(log(income) ~ type + prestige + women, data=Prestige)
summary(reg3)
```

I see now that my adjusted R^2 value increased from 0.7842 to 0.7863. This in interesting in terms of model simplification. I would like to run an ANOVA between the linear model with education and the new linear model without education to see if there is a signficiant difference between the two.

```{r}
anova(reg2, reg3)
```

After running the ANOVA I notice that there is no significant difference between the two models as the Pr(>F) is greater than 0.05. Because of this, I choose not to drop education.

One last analysis I would like to see is the redidual plots of my finale regression model explaining log(income) as the dependent variable.
```{r}
residualPlots(reg2)
```

I see that the residual plots for my regression are very linear and I'm satisfied with the outcome of my model. I choose to not do any further analysis based on these findings.
