---
title: "Lab 6 Income as Dependent"
author: "Kyle Landolt"
date: "April 11, 2015"
output: html_document
---


# Homework
Your homework assignment is to analyze the prestige data, using "income" as the dependent variable.  That is, develop a regression model that helps to explain the income in an occupation, using any of the other data in the the Prestige data set as a potential predictor.  Describe your methodology from beginning to end.  I have done this using Prestige as the dependent variable, and you should follow along with that before continuing.

```{r}
library(spdep)
library(dplyr)
library(ggplot2)
library(GGally)
library(car)
data(Prestige)
```

```{r}
ggpairs(Prestige %>% filter(!is.na(type)) %>% select(education, prestige, women, type, income))
```

I started to analyze the relationship of income as a dependent factor by using the ggpairs function to quickly observe the bivariate relationship of income to the other independent factors. I note the following:
1. If the solid, smoothed black line of income represents its distribution, then I notice that it seems to be positively skewed. 
2. There doesn't seem to a good linear relationship of income to any covariates, although the correlation coefficient for income to prestige was fairly high at +0.703. Education has a relatively high correlation coefficient too at 0.574. It's also noted that income to women have a negative correlation coefficent, possibly showing that women are paid less.
3. The relationships of income to education and prestige both appear exponential.


I do not want to remove any covariates at the moment for risk of creating a confounding variable, so I will do a multivariate regression to see the results.
```{r}
reg0 <- lm(income ~ type + education + prestige + women, data=Prestige)
summary(reg0)
```

The initial multivariate regression has prestige and women as significant relationships. I can check for lineary using the component + residual plots.
```{r}
crPlots(reg0)
```

After checking the plots I noticed that it's still clear that education and prestige have an exponential relationship. Because I don't want to transform my dependent variable first, I will first try to transform education and prestige to exp(education) and exp(prestige). I do this so I won't have to directly transform income in hopes of maintaning my significant relationship of income to women. 
```{r}
reg1 <- lm(income ~ type + exp(education) + exp(prestige) + women, data=Prestige)
summary(reg1)
crPlots(reg1)
```

Transforming exp(educaitona) and exp(prestige) put the data into a nonsensical 

```{r}
reg2 <- lm(log(income) ~ type + education + prestige + women, data=Prestige)
summary(reg2)
crPlots(reg2)
```

